---
title: "Final Project"
author: "Mara VanSickle"
date: "December 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
datedata <- read.csv("Speed Dating Data.csv")
summary(datedata)

#Data for match making accuracy:
#--->(Everything before variable satis_2)


#Variables to change to factors:
datedata$field_cd <- as.factor(datedata$field_cd)
datedata$race <- as.factor(datedata$race)
datedata$gender <- as.factor(datedata$gender)
datedata$goal <- as.factor(datedata$goal)
datedata$date <- as.factor(datedata$date)
datedata$go_out <- as.factor(datedata$go_out)
datedata$career_c <- as.factor(datedata$career_c)
datedata$length <- as.factor(datedata$length)
datedata$numdat_2 <- as.factor(datedata$numdat_2)
datedata$date_3 <- as.factor(datedata$date_3)

#Variables to change to numbers
datedata$income <- as.numeric(datedata$income)
datedata$mn_sat <- as.numeric(datedata$mn_sat)
datedata$tuition <- as.numeric(datedata$tuition)

#Variables to change NA to zero


#Variables to exclude


```


###Questions to Address
Main Question
*Can you predict matches? And which method of doing so is most accurate?
      - Which variables have?

(Additional Questions)
* Who shows up to speed dating?
      - Gender
      - Age
      - Race
      - Income
      - Hobbies
      - Field of Study
      - Zip Code
      - Undergrad vs. Grad
      - etc...
      
* Surveys
      - Comparing surveys over time (which variable matter more short vs. long term)
      - Most and least desirable factors for each gender
      - Visualize survey answers by different demographics/questions

* Who is most likely to have a match?
      - What was the result of those individual surveys?
      - Do people who think more highly of themselves get or give more matches?
      
* How do survey results differ from actual decisions (what they say they like vs. what they actually like)

* Success of matches. Who ended up going on more dates vs. didn't? Which characteristics play a role?
      - Ex = If there is agreement between what you say you like and actually like, do you date more often?


##Getting to know our data
```{r}
#Barplot for Match vs. No Match

nomatch <- nrow(subset(datedata, datedata$match == 0))

##Count how many matches there are
match <- length(which(datedata$match == 1))
match

###Count how many pairs didn't end in match
nomatch <- length(which(datedata$match == 0))
nomatch

match.comparison <- c(match, nomatch)

###Let's graphically display the difference between how many matches and no matches of all pairings created.
barplot(match.comparison, main = "Match vs. No Match", ylab = "Count", col=c("green", "red"))

#Create a new dataset recording the person's preference and demographical background#
date1<-datedata[,c(1,3,11:75,82:86,88:98)]
#Create a new dataset recording the person's partner's demographical background#
date2<-datedata[,c(1,3,11:12,25:75,82:86,88:98)]
date3<-merge(date1, date2, by.x=c('iid','pid'), by.y=c('pid','iid'))
date3<-date3[date3$iid>date3$pid,]
names(date3)
usedate<-date3[,c(5:26,28,30:34,37:40,42:83,97:103,106:152)]

usedate$match<-as.factor(usedate$match)
usedate$career_c.x<-as.factor(usedate$career_c.x)
usedate$career_C.y<-as.factor(usedate$career_c.y)
str(usedate)
names(usedate)

for(i in 1:ncol(usedate[,c(2:22,24:25,27:29,34:73,77:79,85:128)])){
  usedate[is.na(usedate[,i]), i] <- mean(usedate[,i], na.rm = TRUE)
}
names(usedate)
usedate$match<-as.factor(usedate$match)
date.expanded<-as.data.frame(model.matrix(~.-1,data=usedate))
date.expanded<-as.numeric(date.expanded)
normalize <- function(x) {return ((x - min(x)) / (max(x) - min(x)))}
date_n <- as.data.frame(lapply(date.expanded[0:1273], normalize))
#Random Forest to identify key contributers.
library(randomForest)
set.seed(300)
options(na.action='na.pass')
use_clean<-na.omit(usedate)
names(usedate)
```

```{r random forest}
#Random forest to determine decision
names(usedate)
library(caret)
ctrl <- trainControl(method = "repeatedcv", number = 3, repeats = 3)
grid_rf <- expand.grid(.mtry = c(3, 9, 18,36))
usedate$dec.x<-as.factor(usedate$dec.x)
set.seed(300)
m_rf_d <- train(dec.x~int_corr + samerace + age_o + race_o + pf_o_att + pf_o_sin + pf_o_int + pf_o_fun + pf_o_amb + pf_o_sha +  age.x + field_cd.x + mn_sat.x + tuition.x + race.x + imprace.x + imprelig.x + income.x + goal.x + date.x + career_c.x + sports.x + tvsports.x + exercise.x + dining.x + museums.x + art.x + hiking.x + gaming.x + clubbing.x +reading.x + tv.x + theater.x + movies.x + concerts.x + music.x + shopping.x + yoga.x + exphappy.x + expnum.x + + attr1_1.x + 
    sinc1_1.x + intel1_1.x + fun1_1.x + amb1_1.x + shar1_1.x + 
    attr2_1.x + sinc2_1.x + intel2_1.x + fun2_1.x + amb2_1.x + 
    attr3_1.x + sinc3_1.x + fun3_1.x + intel3_1.x + amb3_1.x + 
    attr5_1.x + sinc5_1.x + intel5_1.x + fun5_1.x + amb5_1.x + field_cd.y + mn_sat.y + tuition.y + imprace.y + imprelig.y + income.y + goal.y + date.y + go_out.y + career_c.y + sports.y + tvsports.y + exercise.y + dining.y + museums.y + art.y + hiking.y + gaming.y + clubbing.y + reading.y + tv.y + theater.y + movies.y + concerts.y + music.y + shopping.y + yoga.y + exphappy.y + expnum.y + 
    attr2_1.y + sinc2_1.y + intel2_1.y + fun2_1.y + amb2_1.y + 
    attr3_1.y + sinc3_1.y + fun3_1.y + intel3_1.y + amb3_1.y + 
    attr5_1.y + sinc5_1.y + intel5_1.y + fun5_1.y + amb5_1.y, data = usedate, na.action = 'na.exclude', method = "rf", metric = "Kappa", trControl = ctrl, tuneGrid = grid_rf)
#The final value used for the model was mtry = 3. The Accuracy is 0.719, Kappa statistic is 0.419, which is pretty good for the a rf model#
m_rf_d

#Showing the importance of each variables#
contribution_rf_d<-as.data.frame(importance(m_rf_d$finalModel))
library(data.table)
setDT(contribution_rf_d, keep.rownames = TRUE)[]
contribution_rf_high_d<-contribution_rf[contribution_rf_d$MeanDecreaseGini>10.1,]

#Present the importance of vairable in bar chart#

ggplot(contribution_rf_high_d, aes(x=rn,y=MeanDecreaseGini))+geom_bar(stat="identity")+
     geom_bar(data=subset(contribution_rf_high_d, MeanDecreaseGini==max(MeanDecreaseGini)), aes(rn, MeanDecreaseGini),
              fill="green", stat="identity")


```

```{r random forest}
#Random forest to determine decision of the partner
names(usedate)
library(caret)
ctrl <- trainControl(method = "repeatedcv", number = 3, repeats = 3)
grid_rf <- expand.grid(.mtry = c(3, 9, 18,36))
usedate$dec.y<-as.factor(usedate$dec.y)
set.seed(300)
m_rf_d0 <- train(dec.y~int_corr + samerace + age_o + race_o + pf_o_att + pf_o_sin + pf_o_int + pf_o_fun + pf_o_amb + pf_o_sha +  age.x + field_cd.x + mn_sat.x + tuition.x + race.x + imprace.x + imprelig.x + income.x + goal.x + date.x + career_c.x + sports.x + tvsports.x + exercise.x + dining.x + museums.x + art.x + hiking.x + gaming.x + clubbing.x +reading.x + tv.x + theater.x + movies.x + concerts.x + music.x + shopping.x + yoga.x + exphappy.x + expnum.x + + attr1_1.x + 
    sinc1_1.x + intel1_1.x + fun1_1.x + amb1_1.x + shar1_1.x + 
    attr2_1.x + sinc2_1.x + intel2_1.x + fun2_1.x + amb2_1.x + 
    attr3_1.x + sinc3_1.x + fun3_1.x + intel3_1.x + amb3_1.x + 
    attr5_1.x + sinc5_1.x + intel5_1.x + fun5_1.x + amb5_1.x + field_cd.y + mn_sat.y + tuition.y + imprace.y + imprelig.y + income.y + goal.y + date.y + go_out.y + career_c.y + sports.y + tvsports.y + exercise.y + dining.y + museums.y + art.y + hiking.y + gaming.y + clubbing.y + reading.y + tv.y + theater.y + movies.y + concerts.y + music.y + shopping.y + yoga.y + exphappy.y + expnum.y + 
    attr2_1.y + sinc2_1.y + intel2_1.y + fun2_1.y + amb2_1.y + 
    attr3_1.y + sinc3_1.y + fun3_1.y + intel3_1.y + amb3_1.y + 
    attr5_1.y + sinc5_1.y + intel5_1.y + fun5_1.y + amb5_1.y, data = usedate, na.action = 'na.exclude', method = "rf", metric = "Kappa", trControl = ctrl, tuneGrid = grid_rf)
#The final value used for the model was mtry = 36. The Accuracy is 0.719, Kappa statistic is 0.371, which is pretty good for the a rf model#
m_rf_d0

#Showing the importance of each variables#
contribution_rf_d0<-as.data.frame(importance(m_rf_d0$finalModel))
library(data.table)
setDT(contribution_rf_d0, keep.rownames = TRUE)[]
contribution_rf_high_d0<-contribution_rf_d0[contribution_rf_d0$MeanDecreaseGini>14,]

#Present the importance of vairable in bar chart#

ggplot(contribution_rf_high_d0, aes(x=rn,y=MeanDecreaseGini))+geom_bar(stat="identity")+
     geom_bar(data=subset(contribution_rf_high_d0, MeanDecreaseGini==max(MeanDecreaseGini)), aes(rn, MeanDecreaseGini),
              fill="green", stat="identity")


```


```{r random forest to predict match}

library(caret)
f<-as.formula(paste("match~",paste(names(usedate),collapse = "+")))
f
ctrl <- trainControl(method = "repeatedcv", number = 3, repeats = 3)
usedate$match<-as.factor(usedate$match)
grid_rf <- expand.grid(.mtry = c(3, 9, 18, 36))
set.seed(300)
m_rf <- train(match~int_corr + samerace + age_o + race_o + pf_o_att + pf_o_sin + pf_o_int + pf_o_fun + pf_o_amb + pf_o_sha + age.x + field_cd.x + mn_sat.x + tuition.x + race.x + imprace.x + imprelig.x + income.x + goal.x + date.x + career_c.x + sports.x + tvsports.x + exercise.x + dining.x + museums.x + art.x + hiking.x + gaming.x + clubbing.x +reading.x + tv.x + theater.x + movies.x + concerts.x + music.x + shopping.x + yoga.x + exphappy.x + expnum.x + + attr1_1.x + 
    sinc1_1.x + intel1_1.x + fun1_1.x + amb1_1.x + shar1_1.x + 
    attr2_1.x + sinc2_1.x + intel2_1.x + fun2_1.x + amb2_1.x + 
    attr3_1.x + sinc3_1.x + fun3_1.x + intel3_1.x + amb3_1.x + 
    attr5_1.x + sinc5_1.x + intel5_1.x + fun5_1.x + amb5_1.x + + field_cd.y + mn_sat.y + tuition.y + imprace.y + imprelig.y + income.y + goal.y + date.y + go_out.y + career_c.y + sports.y + tvsports.y + exercise.y + dining.y + museums.y + art.y + hiking.y + gaming.y + clubbing.y + reading.y + tv.y + theater.y + movies.y + concerts.y + music.y + shopping.y + yoga.y + exphappy.y + expnum.y + 
    attr2_1.y + sinc2_1.y + intel2_1.y + fun2_1.y + amb2_1.y + 
    attr3_1.y + sinc3_1.y + fun3_1.y + intel3_1.y + amb3_1.y + 
    attr5_1.y + sinc5_1.y + intel5_1.y + fun5_1.y + amb5_1.y , data = usedate, na.action = 'na.exclude', method = "rf", metric = "Kappa", trControl = ctrl, tuneGrid = grid_rf)

#The final value used for the model was mtry = 36. The accuracy is 0.843,which is relatively high#
m_rf_d0$finalModel
#Showing the importance of each variables#
contribution_rf_d0<-as.data.frame(importance(m_rf_d0$finalModel))
library(data.table)
setDT(contribution_rf_d0, keep.rownames = TRUE)[]
contribution_rf_high_d0<-contribution_rf_d0[contribution_rf_d0$MeanDecreaseGini>7.5,]

#Present the importance of vairable in bar chart#

ggplot(contribution_rf_high, aes(x=rn,y=MeanDecreaseGini))+geom_bar(stat="identity")+
     geom_bar(data=subset(contribution_rf_high_d0, MeanDecreaseGini==max(MeanDecreaseGini)), aes(rn, MeanDecreaseGini),
              fill="green", stat="identity")


```

```{r logit models}
#Logit models to identify key drivers for match options
library(gmodels)
logit.model1 <- glm(match ~ int_corr + samerace + age_o + pf_o_att +
+ pf_o_sin + pf_o_int + pf_o_fun + pf_o_amb + pf_o_sha + attr_o.x + sinc_o.x + intel_o.x + fun_o.x + amb_o.x + shar_o.x , data = usedate, family = "binomial",na.action = 'na.exclude')

summary(logit.model1)
names(usedate)
useinterest<-usedate
```



