---
title: "Final Project"
author: "Mara VanSickle"
date: "December 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
datedata <- read.csv("Speed Dating Data.csv")
summary(datedata)

#Data for match making accuracy:
#--->(Everything before variable satis_2)


#Variables to change to factors:
field_cd <- as.factor(field_cd)
race <- as.factor(race)
gender <- as.factor(gender)
goal <- as.factor(goal)
date <- as.factor(date)
go_out <- as.factor(go_out)
career_c <- as.factor(career_c)
length <- as.factor(length)
numdat_2 <- as.factor(numdat_2)
date_3 <- as.factor(date_3)

#Variables to change to numbers
income <- as.numeric(income)
mn_sat <- as.numeric(mn_sat)
tuition <- as.numeric(tuition)

#Variables to change NA to zero


#Variables to exclude


```


###Questions to Address
Main Question
*Can you predict matches? And which method of doing so is most accurate?
      - Which variables have?

(Additional Questions)
* Who shows up to speed dating?
      - Gender
      - Age
      - Race
      - Income
      - Hobbies
      - Field of Study
      - Zip Code
      - Undergrad vs. Grad
      - etc...
      
* Surveys
      - Comparing surveys over time (which variable matter more short vs. long term)
      - Most and least desirable factors for each gender
      - Visualize survey answers by different demographics/questions

* Who is most likely to have a match?
      - What was the result of those individual surveys?
      - Do people who think more highly of themselves get or give more matches?
      
* How do survey results differ from actual decisions (what they say they like vs. what they actually like)

* Success of matches. Who ended up going on more dates vs. didn't? Which characteristics play a role?
      - Ex = If there is agreement between what you say you like and actually like, do you date more often?


##Getting to know our data
```{r}
#Barplot for Match vs. No Match

nomatch <- nrow(subset(datedata, datedata$match == 0))

##Count how many matches there are
match <- length(which(datedata$match == 1))
match

###Count how many pairs didn't end in match
nomatch <- length(which(datedata$match == 0))
nomatch

match.comparison <- c(match, nomatch)

###Let's graphically display the difference between how many matches and no matches of all pairings created.
barplot(match.comparison, main = "Match vs. No Match", ylab = "Count", col=c("green", "red"))












```



